---
title: Simulating exponential survival times
output:
  html_document:
    code_download: true
---

We're going to try simulating survival data with exponentially distributed failure times. Bunch of ways to do 
this:

- Using the `Distributions.jl` package to sample exponential random variates. 
- Using the inverse CDF method, i.e., sampling Unif(0,1) variates and using the survival function (Crowther and Lambert).
- By integrating an exponential density to get the exponential CDF and then appealing to the previous method.
- Other algorithms, e.g., uniformization, importance/rejection sampling, MCMC, etc.

notes from 25th May 2021:

### Environment

Julia can be run from JUPYTER notebooks ('JU')

Install JULIA (https://julialang.org/downloads/) and install packages for JULIA JUPYTER: https://datatofish.com/add-julia-to-jupyter/

I ran this from the Julia REPL but there are many other options...
import Pkg
using Pkg
Pkg.add("IJulia")

Then start a JUPYTER notebook and select language -> Julia (version)
Example JUPYTER notebook include
There are some differences in code (eg: scoping) between JUPYTER and Rstudio (detailed in code)

Julia can be run from Rstudio in Rnotebook
https://github.com/Non-Contradiction/JuliaCall

In Rstudio: 
```{r}
#install.packages("JuliaCall")
library(JuliaCall)
julia_setup(JULIA_HOME = 'C:/Users/hartwickma/AppData/Local/Programs/Julia-1.6.1/bin/') #specific to where your Julia installed
julia <- julia_setup()
library(tidyverse)

#plot theme
mytheme <- theme(
  panel.border = element_blank(),
  panel.grid.major = element_line(color = 'grey60', linetype = 'dashed',
                                  size = 0.5),
  panel.grid.minor = element_blank(),
  panel.background = element_rect(colour = "white", fill = "white"),
  legend.position = "right", axis.line = element_line(colour = "white"),
  legend.key = element_rect(fill = NA),
  legend.text.align = 0,
  text = element_text(color = 'grey5'),
  axis.text.x = element_text(colour = "grey5", size = 10),
  axis.text.y = element_text(colour = "grey5", size = 10,),
  axis.ticks = element_line(colour = "grey5"),
  title = element_text(colour = 'grey5', size = 10))
```

Now you're ready to run Julia from JUPYTER notebook or Rnotebook (Rmarkdown works too...different discussion)

Visual studio documentation: ?


### Rstudio Rnotebook

quick Rnotebook shortcuts: 
ctrl + alt + I to insert code chunks
ctrl + shitft + enter to execute code chunk

test that julia can be found in rstudio

```{julia}

println("Hello world (in Julia from an Rnotebook...)")

```
#### set up environment to work with distributions and plotting


```{julia}
import Pkg
Pkg.add("Distributions")
Pkg.add("Random")
Pkg.add("Plots")
Pkg.add("GR")
Pkg.add("DataFrames")

```

#### background
In R:
p for "probability", the cumulative distribution function (c. d. f.)
q for "quantile", the inverse c. d. f.
d for "density", the density function (p. f. or p. d. f.)
r for "random", a random variable having the specified distribution


Distribution	        Functions
Beta	               pbeta	qbeta	dbeta	rbeta
Binomial	           pbinom	qbinom	dbinom	rbinom
Cauchy	             pcauchy	qcauchy	dcauchy	rcauchy
Chi-Square	         pchisq	qchisq	dchisq	rchisq
Exponential	         pexp	qexp	dexp	rexp
F	                   pf	qf	df	rf
Gamma	               pgamma	qgamma	dgamma	rgamma
Geometric	           pgeom	qgeom	dgeom	rgeom
Hypergeometric	     phyper	qhyper	dhyper	rhyper
Logistic	           plogis	qlogis	dlogis	rlogis
Log Normal	         plnorm	qlnorm	dlnorm	rlnorm
Negative Binomial	   pnbinom	qnbinom	dnbinom	rnbinom
Normal	             pnorm	qnorm	dnorm	rnorm
Poisson	             ppois	qpois	dpois	rpois
Student t	           pt	qt	dt	rt
Studentized Range	   ptukey	qtukey	dtukey	rtukey
Uniform	             punif	qunif	dunif	runif
Weibull	             pweibull	qweibull	dweibull	rweibull
Wilcoxon Rank Sum Statistic	pwilcox	qwilcox	dwilcox	rwilcox
Wilcoxon Signed Rank Statistic	psignrank	qsignrank	dsignrank	rsignrank


generate n random numbers from a normal distribution with given mean & st. dev.
http://uc-r.github.io/generating_random_numbers/#normal
rnorm(n, mean = 0, sd = 1)    

generate CDF probabilities for value(s) in vector q 
pnorm(q, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)    

generate quantile for probabilities in vector p (inverse cdf)
qnorm(p, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)    

generate density function probabilites for value(s) in vector x
dnorm(x, mean = 0, sd = 1)  

```{r}
set.seed(600)
x <- rnorm(100, mean = 0, sd = 1)
x
x %>%
  as_tibble() %>%
  ggplot(., aes(value)) +
  geom_histogram(bins = 10) +
  mytheme
```


```{r}
p_x <- pnorm(x, mean = 0, sd = 1)
t <- cbind(X = x, cdf = p_x)
head(t)
t %>%
  as_tibble() %>%
  ggplot(., aes(x=X, y = cdf)) +
  geom_line() +
  mytheme

```


```{r}
# usage: the probability of a value given mean and sd
dnorm(27.4, mean=0, sd=1)
# or
d_x <- dnorm(x, mean = 0, sd = 1)
t <- cbind(t, pdf = d_x)
head(t)
t %>%
  as_tibble() %>%
  ggplot(., aes(x=X, y = pdf)) +
  geom_line() +
  mytheme
```


```{r}
#returns the quantile with the specified probability given mean and st  
#inverse cdf
i <- qnorm(0.95, mean=0, sd=1)
label <- print(paste('95% of values in x', '\n', 'are less than', round(i,3)))

t %>%
  as_tibble() %>%
  ggplot(., aes(x=X)) +
  geom_line(aes(y = cdf), color = 'green') +
  geom_line(aes(y = pdf), color = 'skyblue1') +
  geom_vline(xintercept = i, color = 'red', linetype = 2) +
  annotate('text', label = label, x = 1, y = 0.5, color = 'grey10') +
  ylab('probability') +
  mytheme

```


```{r}
#so...create a vector of quantiles given their probability (inverse cdf)
t <- t %>%
  as_tibble() %>%
  mutate(i_cdf = qnorm(pdf, mean=0, sd=1))

t %>%
  as_tibble() %>%
  ggplot(., aes(x=X)) +
  geom_col(aes(y= i_cdf), color = 'red', linetype = 2) +
  mytheme
```

So in exponential distribution:

generate a vector of length n with rate = 1
rexp(n, rate = 1)   

generate CDF probabilities for value(s) in vector q when rate = 4.
pexp(q, rate = 1)   

generate quantile for probabilities in vector p when rate = 4.
qexp(p, rate = 1)   

generate density function probabilites for value(s) in vector x when rate = 4.
p <- dexp(x, rate = 1) 

#### To simulate failure time given a probabilty (create x):

```{r}

survival_df <- function(N, distribution,...){
#create a vector of survival times for n individuals choice of distribution

#step one: seq of individuals
individual <-seq(1:N) #

#distribution (normal, exp, wilcox)
 x <- if (distribution == 'exponential' | distribution == 'exp') {rexp(N)
                                  } else if (distribution == 'wilcox') {rwilcox(N)
                                  } else if (distribution == 'normal') {rnorm(N)
                                  }

pdf <- numeric()
for(i in x){pdf <- c(pdf, if(distribution == 'exponential' | distribution == 'exp') {dexp(i)
                                  } else if (distribution == 'wilcox') {dwilcox(i)
                                  } else if (distribution == 'normal') {dnorm(i)
                                  })
}


i_cdf = numeric()
for(i in pdf){i_cdf <- c(i_cdf, if(distribution == 'exponential' | distribution == 'exp') {qexp(i)
                                  } else if (distribution == 'wilcox') {qwilcox(i)
                                  } else if (distribution == 'normal') {qnorm(i)
                                  })
}


df <- cbind(individual,x, pdf, i_cdf)
df<-as.data.frame(df)
return(df)
}

```

```{r}
survival_df(N = 15, distribution = 'exp') %>% 
  ggplot(., aes(x = x)) + 
  geom_line(aes(y = i_cdf)) +
  geom_col(aes(y = i_cdf, fill = as.character(individual))) +
  mytheme

```
```{r}
z <- 1-pexp(2)
print(paste('given the sample, probability of survival past 2 is less than', round(z,2)))
```

### In Julia

```{julia}

#q - what does '!' do here?

#resources: https://www.youtube.com/watch?v=vUA2NzgzH_g

using Distributions
using Random
Random.seed!(1234)
rand(2)
#sequences
#The collect() is an inbuilt function in julia which is used to return an array of all items in the specified collection or iterator.
#you can use the range() function, which allows you to specify the length, step size, or both
collect(range(0, stop = 5, length = 3))

#logccdf(d::UnivariateDistribution, x::Real)

```

```{julia}
#generate a sequence of numbers
collect(range(0, stop = 5, length = 3) )
```

```{julia}
randn(10)
```
```{julia}
#oops accidentally overwrote randexp
randexp = 10
```

```{julia}
#functionality can be restored by resetting, use the corresponding package or 'Base' preceeding
#https://stackoverflow.com/questions/65902105/how-to-reset-any-function-in-julia-to-its-original-state
randexp = Random.randexp
randexp(10)
```

```{julia}

# generate n random numbers from a normal distribution with given mean & st. dev.
# for norm and exp, there are built ins
Random.randn(10)   #normal distribution
Random.randexp(10)    #exponential distribution
```

https://juliastats.org/Distributions.jl/v0.14/univariate.html

Arcsine(a, b)
Beta(ð›¼, ð›½)
BetaPrime(ð›¼, ð›½)
Cauchy(ðœ‡, ðœŽ)
Chi(ðœˆ)
Chisq(ðœˆ)
Erlang(ð›¼, ðœƒ)
Exponential(b) # Exponential distribution with scale b
FDist(ðœˆ1, ðœˆ2)
Frechet(ð›¼, ðœƒ)
Gamma(ð›¼, ðœƒ)
GeneralizedExtremeValue(ðœ‡, ðœŽ, ðœ‰)
Gumbel() # Gumbel distribution with zero location and unit scale, i.e. Gumbel(0.0, 1.0)
InverseGamma() # Inverse Gamma distribution with unit shape and unit scale, i.e. InverseGamma(1.0, 1.0)
InverseGaussian() # Inverse Gaussian distribution with unit mean and unit shape, i.e. InverseGaussian(1.0, 1.0)
Laplace() # Laplace distribution with zero location and unit scale, i.e. Laplace(0.0, 1.0)
Levy() # Levy distribution with zero location and unit scale, i.e. Levy(0.0, 1.0)
LogNormal(ðœ‡, ðœŽ)
Logistic(ðœ‡, ðœƒ)
Normal() # standard Normal distribution with zero mean and unit variance
Pareto() # Pareto distribution with unit shape and unit scale, i.e. Pareto(1.0, 1.0)
TDist(d) # t-distribution with d degrees of freedom
TriangularDist(a, b) # Triangular distribution with lower limit a, upper limit b, and mode (a+b)/2
VonMises(ðœ‡, ðœ…)
Weibull(ð›¼, ðœƒ) #The Weibull distribution with shape ð›¼ and scale ðœƒ has probability density function
Bernoulli(p) #Bernoulli distribution is parameterized by a success rate p, probability 1-p.
Binomial(n, p) #A Binomial distribution characterizes the number of successes: n, the number of trials, and p, the probability of success in an individual trial, with the distribution:
Categorical(p) # Categorical distribution is parameterized by a probability vector p (of length K).
DiscreteUniform(a, b) #Discrete uniform distribution is over a consecutive sequence of integers between a and b, inclusive.
Geometric(p) #A Geometric distribution characterizes the number of failures before the first success in a sequence of independent Bernoulli trials with success rate p.
Hypergeometric(s, f, n) #A Hypergeometric distribution describes the number of successes in n draws without replacement from a finite
        #population containing s successes and f failures.
NegativeBinomial(r, p) # Negative binomial distribution with r successes and success rate p
Poisson(lambda) # Poisson distribution with rate parameter lambda
PoissonBinomial(p) # Poisson Binomial distribution with success rate vector p
Skellam(mu1, mu2) # Skellam distribution for the differen

```{julia}

# also how in the Distributions package

d = Distributions.Normal()
x = Random.rand(d, 100)
```


```{julia}
x = Random.rand(Distributions.Normal(), 100)
```


```{julia}
#equivalents
#https://readthedocs.org/projects/blopmatching/downloads/pdf/latest/
pdf 
logpdf
cdf
logcdf
ccdf #The complementary cumulative function evaluated at x, i.e. 1 - cdf(d, x).
logccdf
quantile
cquantile
invlogcdf
invlogccdf
```

```{julia}
using Plots
histogram(x)

```

julia equivalents
https://arxiv.org/pdf/1907.08611.pdf
R                   Julia
rnorm(n, mu, sig) = rand(Normal(mu, sig), n)
dnorm(x, mu, sig) = pdf(Normal(mu, sig), x)
pnorm(x, mu, sig) = cdf(Normal(mu, sig), x)
qnorm(p, mu, sig) = quantile(Normal(mu, sig), p)


```{julia}
#Pkg.add("StatsFuns") vs Distributions
#takes the specified outside of the wrapper but doesnt have all the distributions
#https://github.com/JuliaStats/StatsFuns.jl

using StatsFuns
x=rand(Normal(),100)
y=normpdf.(x)
z=pdf.(Normal(), x)
plot(x,z, seriestype = bar)
```


```{julia}

using GR

# a = rand(Normal(mu, sig), n)
# b = pdf(Normal(mu, sig), x)
# c = cdf(Normal(mu, sig), x)
# d = ccdf(Normal(),p)
# e = quantile(Normal(mu, sig), p)

a = rand(Normal(), 120)
b = pdf(Normal(), a)
c = cdf(Normal(), a)
#d = ccdf(Normal(),b)
e = quantile(Normal(), 0.95)

#http://docs.juliaplots.org/latest/generated/supported/ 
plot(a, b, seriestype = :line, label = "pdf")
plot!(a, c, seriestype = :line, label = "cdf")
plot!([e], seriestype="vline", label = "i_cdf")
# t %>%
#   as_tibble() %>%
#   ggplot(., aes(x=X)) +
#   geom_line(aes(y = cdf), color = 'green') +
#   geom_line(aes(y = pdf), color = 'skyblue1') +
#   geom_vline(xintercept = i, color = 'red', linetype = 2) +
#   annotate('text', label = label, x = 1, y = 0.5, color = 'grey10') +
#   ylab('probability')
#   mytheme
```


```{julia}

#create a function that wraps cdf etc around the N and plot

function f(x,y)
           x + y
       end
       
f(2,3)
```

```{julia}
using DataFrames

function survival_julia(N, dist)
  individual = collect(1:N)
  a = rand(dist, N)
  b = pdf(dist, a)
  c = quantile(dist, b)
  return df = DataFrame(individual = individual,
                        x = a,
                        pdf = b,
                        i_cdf = c)
  end
#create DF with any distribution, rate can be specified when appropriate
s_j = survival_julia(10, Exponential(2))

plot(s_j[:,2], s_j[:,4], seriestype = :line)
#plot!(s_j[,2], s_j[,], seriestype = :bar, label = "cdf")
```
```{r}
#F^(-1)(u)=(-log(1-u))/Î¸
#where u is selected from a uniform distribution between 0 and 1 and theta is the 'rate' (angle)
set.seed(1)
exp_icdf <- function(n, theta) {
  u<-runif(n)
  i_cdf_new<-(-log(1 - u))/ theta
  i_cdf_wheel <- qexp(u, rate = theta)
  df <- as.data.frame(cbind(u, i_cdf_new, i_cdf_wheel))
  return(df)
}

exp_icdf(15, 2) -> x
x
```

```{julia }
Random.seed!(1)
using Random

rand(15)
```


```{julia}

#the rand function in julia creates a uniform distribution between 0 and 1 by default - range can be specified as can n
#https://www.roelpeters.be/sampling-from-a-specific-distribution-in-julia/

using Random

function icdfExp(u,theta)
  x = (-log(1 - u))/ theta
end

a = rand(15)
d = []
for i in a
  #println(icdfExp(i,2))
  push!(d,icdfExp(i,2))
end

c = quantile(Exponential(2), a)
df = DataFrame(a=a, d=d, c=c)
```

```{julia}
#Distribution calc for quantile?
#https://github.com/JuliaStats/Distributions.jl/blob/master/src/univariate/continuous/exponential.jl
#-xval(d, log1p(-p))

```


```{r}
#https://rdrr.io/cran/cpsurvsim/src/R/weibullfns.R
set.seed(1)

#this must be the discrete i_cdf?

# weib_icdf <- function(n, gamma, theta) {
#   if(is.numeric(n) == FALSE | n <= 0) {
#     stop("n must be an integer greater than 0.")
#   }
#   if(is.numeric(theta) == FALSE | is.numeric(gamma) == FALSE |
#      theta <= 0 | gamma <= 0) {
#     stop("Theta and gamma must be numeric and > 0.")
#   }
#   x <- runif(n)
#   return(((gamma / theta) * x) ^ (1 / gamma))
# }

#from wikipedia these are continous then?
#{\displaystyle Q(p;k,\lambda )=lambda(-ln(1-p))^{1/k}}
#matlab âˆ’a[ln(1âˆ’p)]^1/b
```


```{r}
theta <- 1
gamma <- 2
n <- 4

u<-runif(n)
i_cdf_new<-(theta*(-log(1-u))) ^(1/gamma)
```

```{r}
#gamma as shape parameter, theta as rate(scale?) parameter, n draw from uniform distribution between 0 and 1
replicatedWeibIcdf <- function(n, gamma, theta) {
  set.seed(1)
  u<-runif(n)
  i_cdf_new<-(theta*(-log(1-u))) ^ (1 / gamma)
  i_cdf_wheel <- qweibull(p = u, shape = gamma, scale = theta)
  df <- as.data.frame(cbind(u, i_cdf_new, i_cdf_wheel))
  return(df)
}

```

```{julia}

#problem with seeds producing different sequences between R and Julia but when same numbers are input our Julia function produces same outcomes as R (our function and quantile function)

using Random
Random.seed!(1)

function icdfWeib(u,theta, gamma)
  x = (theta*(-log(1-u))) ^ (1 / gamma)
end

#a = rand(4)
a = [0.2655087, 0.3721239, 0.5728534, 0.9082078]
d = []
for i in a
  #println(icdfExp(i,2))
  push!(d,icdfWeib(i,1,2))
end

c = quantile(Exponential(2), a) # ignore
df = DataFrame(a=a, d=d, c=c)

```
```{r}

#qgamma(p = u, shape = shape, rate = rate)

rGammaicdf <- function(n, shape) {
  set.seed(1)
  u<-runif(n)
  i_cdf_new<-(1/theta*(-log(u)))
  i_cdf_wheel <- qgamma(u, shape = shape)
  df <- as.data.frame(cbind(u, i_cdf_new, i_cdf_wheel))
  return(df)
}

```


```{julia}
#https://wiki.math.uwaterloo.ca/statwiki/index.php?title=techniques_for_Normal_and_Gamma_Sampling
#finding this equation was ...fun? xi=âˆ’1/Î»(log(ui)

```

